Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2015-09-23 notes/paper-degenerate-linear-systems.tex
Proofread and final edit.

\begin{abstract}
Linear Gaussian state space models are ubiquitous in signal processing, and an important procedure is that of estimating system parameters from observed data. Rather than making a single point estimate, it is often desirable to conduct Bayesian learning, in which the entire posterior distribution of the unknown parameters is sought. This can be achieved using Markov chain Monte Carlo. On some occasions it is possible deduce the form of the unknown system matrices in terms of a small number of scalar parameters, by considering the underlying physical processes involved. Here we study the case where this is not possible, and the entire matrices must be treated as unknowns. An efficient Gibbs sampling algorithm exists for the basic formulation of linear model. We extend this to the more challenging situation where the transition model is degenerate. Appropriate Markov kernels are devised and demonstrated with simulations.
\end{abstract}

@@

A linear Gaussian state space model is specified by a number of system matrices which govern the latent state and observation processes. An important consideration is how to learn these fixed matrices from observed data. The parametric approach to this problem is to deduce the form of the system matrices in terms of a small number of scalar parameters by considering the physical mechanisms underlying the system, and then attempt to learn these. (See e.g. \cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned. This type of non-parametric learning for linear Gaussian state space models is the focus of this paper. There is a large body of research on this topic, which has generally concentrated on point estimate methods. These include subspace identification approaches \cite{VanOverschee1991,Viberg1995}, and maximum likelihood estimates using either gradient-based algorithms (see e.g. \cite{Cappe2005,Sarkka2013}) or Expectation-Maximisation (EM) \cite{Shumway1982,Digalakis1993,Ghahramani1996,Gibson2005,Li2009}.

@@

The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models. However, it is not able to handle the case in which the transition model is degenerate, i.e. when the transition covariance matrix is not full rank. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. Furthermore, reversible jump construction \cite{Green1995,Green2009} is introduced which enables the sampler to learn the rank of the transition covariance matrix.

@@

The paper is structured as follows. In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model. In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions, and in section~\ref{sec:rank_learning} we show how the rank of the transition covariance may be learned. Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations}.

@@

The system is parameterised by the four matrices $\left\{\lgtm, \lgom, \lgtv, \lgov\right\}$. In this paper we focus on learning the transition model, i.e. $\lgtm$ and $\lgtv$, and treat $\lgom$ and $\lgov$ as fixed and known throughout. Conditional on these matrices, state inference may be carried out analytically. Posterior filtering and smoothing densities may be calculated using the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. Furthermore, the Kalman filter may also be used to evaluate the marginal likelihood $\den(\ob{1:\timax}|\lgtm, \lgtv)$, and it is possible to draw posterior samples from the state posterior $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$ using the forward-filtering-backward-sampling method \cite{Chib1996}. The purpose of a learning algorithm is estimate unknown values of $\left\{\lgtm, \lgtv\right\}$ from a sequence of observations. Within the Bayesian framework this means calculating or approximating the posterior distribution $\den(\lgtm, \lgtv|\ob{1:\timax})$, which we can achieve using MCMC.

In order to conduct Bayesian learning, we first need to define prior distributions for the unknown system matrices $\lgtm$ and $\lgtv$. Although these should be selected to reflect prior belief about the system in question, there is substantial benefit in using conjugate priors, as these enable us to use Gibbs sampling moves. Typically, we will not have strong prior beliefs about the parameters, so it is reasonable to use an uninformative conjugate prior. However, if we do have prior knowledge to take into account which means that a conjugate prior is not appropriate, then we can treat each Gibbs move as a proposal in a Metropolis-Hastings scheme and use an accept/reject stage to account for the difference between the true prior and the conjugate prior used for the sampling \cite{Wills2012}.

@@

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
\end{IEEEeqnarray}
%
in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$. This is almost equivalent to the previous, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce a parameterisation for such degenerate transition models which can uniquely represent any such system, and show how MCMC moves may be implemented for learning. Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.

The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix. To avoid this ambiguity, it is better to focus on learning $\lgtv$ as before, and to define $\lgdm$ as some matrix square root of $\lgtv$ (e.g. the Cholesky factor or the principal matrix square root). In this new setting, $\lgtv$ is a positive semi-definite matrix with rank $\rk$.

@@

There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach. However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult. We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model. In this section we show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.

@@

We can also write the ``full'' factorisation by finding a set of $\lsd-\rk$ additional orthonormal vectors and appending them to $\tvvec$, along with matching zeros to $\tvval$, as follows,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \begin{bmatrix}\tvvec & \tvvecorth \end{bmatrix} \begin{bmatrix}\tvval & \zmat \\ \zmat & \zmat \end{bmatrix} \begin{bmatrix}\tvvec\tr \\ \tvvecorth\tr \end{bmatrix}
\end{IEEEeqnarray}
%
Note that $\tvvecorth$ is not unique.

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation). For the rank-$\rk$ positive semi-definite covariance matrix $\lgtp$,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
\end{IEEEeqnarray}
%
where,
%
$\tvfull$ is a $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

An appropriate prior distribution for the degenerate transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution to the desired rank, $\priordof=\rk$. This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrixrank{\rk}}     .
\end{IEEEeqnarray}
%
The density (with respect to the Lesbegue measure on the distinct elements) is,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrixrank{\rk}}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
 & & \qquad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrixrank{\rk} \right)    ,
\end{IEEEeqnarray}
%
in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$. A prior estimate of $\lgtv$ is $\frac{1}{\rk}\priorscalematrixrank{\rk}$. Thus, in order to maintain consistency between the different ranks, we should set $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$, where $\priorscalematrixbase$ is an estimate of $\lgtv$.

The matrix normal component of the prior is more problematic. Although the distribution it is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model. (See \cite{Diaz-Garcia2006} for a discussion of the singular matrix normal distribution.)

@@

Using this prior, we need a mechanism for choosing $\priorextratpval$. These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$. A simple choice is to set $\priorextratpval=\priortypval\idmat$ where $\priortypval$ is a prior estimate of a ``typical'' eigenvalue. Alternatively, $\priortypval$ could be made dependent on $\tvval$; for example it could be set to the average (arithmetic or geometric) of the current eigenvalues. In practice, uninformative priors will often be used. If $\priorcolumnvariance\to\infty$, then the effect of $\priortypval$ becomes negligible.

@@

As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$. Nevertheless, we can sample from these distributions exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions. These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.

@@

The likelihood function for the entire state trajectory is thus,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) = \den(\ls{0}) \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
 &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
 && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
 %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
\end{IEEEeqnarray}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

Writing all the subspace constraints concisely in a single equation,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times\lsd}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}    .
\end{IEEEeqnarray}

@@

\section{Reversible Jump for Rank Learning} \label{sec:rank_learning}

@@

First we introduce a prior distribution for the rank of $\lgtv$,
%
\begin{IEEEeqnarray}{rCl}
 \den(\rk) &=& \sum_{i=1}^{\lsd} \priorrankweight{i} \diracden{\rk}{i}     ,
\end{IEEEeqnarray}
%
where $\diracden{j}{i}$ is a probability point mass at $j=i$, and $\sum_i \priorrankweight{i} = 1$. The target distribution is now,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\lgtm, \lgtv, \rk) &\propto& \den(\ob{1:\timax}|\lgtm, \lgtv) \den(\lgtm, \lgtv|\rk) \den(\rk)    .
\end{IEEEeqnarray}

@@

For the sake of simplicity we suggest sampling the new eigenvalue and eigenvector independently from uniform distributions. For the eigenvalue,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) = \begin{cases}
                                          \frac{1}{\eval{\rk}} & 0 < \eval{\rk+1} < \eval{\rk} \\
                                          0 & \text{ otherwise}    .
                                         \end{cases}
\end{IEEEeqnarray}
%
For the eigenvector,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) = \begin{cases}
                                          \frac{1}{C} & \evec{\rk+1} \in \evecset{\rk+1} \\
                                          0 & \text{ otherwise}    .
                                         \end{cases}
\end{IEEEeqnarray}
%
$C$ is the volume of the manifold, which is half of a $(d-r-1)$-sphere (half because of the requirement that the sign of the first element be positive) \cite{Muirhead1982}. Hence,
%
\begin{IEEEeqnarray}{rCl}
 C &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}     .
\end{IEEEeqnarray}
%
In order to generate such a vector, first draw a $(\lsd-\rk)$-dimensional sample from $\normaldist{\zmat}{\idmat}$, normalise it, and then multiply by $\tvvecorth$, which is an arbitrary matrix of orthonormal columns such that $\tvvec\tr\tvvecorth=\zmat$.

The acceptance probability for this move and its reverse respectively are \cite{Green1995},
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& \min\left\{1, \beta \right\} \\
 \mhap(\lgtv\mcnew\to\lgtv,\eval{\rk+1},\evec{\rk+1}) &=& \min\left\{1, \beta\inv \right\} 
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
\end{IEEEeqnarray}
%
where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk+1\to\rk)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.

@@

\section{Discussion and Conclusions}

>>>>

\begin{abstract}
Linear Gaussian state space models are ubiquitous in signal processing, and an important procedure is that of estimating system parameters from observed data. Rather than making a single point estimate, it is often desirable to conduct Bayesian learning, in which the entire posterior distribution of the unknown parameters is sought. This can be achieved using Markov chain Monte Carlo. On some occasions it is possible deduce the form of the unknown system matrices in terms of a small number of scalar parameters, by considering the underlying physical processes involved. Here we study the case where this is not possible, and the entire matrices must be treated as unknowns. An efficient Gibbs sampling algorithm exists for the basic formulation of linear model. We extend this to the more challenging situation where the transition model is possibly degenerate. Appropriate Markov kernels are devised and demonstrated with simulations.
\end{abstract}

@@

A linear Gaussian state space model is specified by a number of system matrices which govern the latent state and observation processes. An important consideration is how to learn these fixed matrices from observed data. Sometimes it is possible to deduce the form of the system matrices in terms of a small number of scalar parameters by considering the physical mechanisms underlying the system, and then attempt to learn these. (See e.g. \cite{Kantas2009,Andrieu2010}.) However, in many systems there may not be an obvious parametric form, and it may be better to treat each system matrix in its entirety as an unknown variable to be learned; this is the case studied in this paper. There is a large body of research on this topic, which has generally concentrated on point estimate methods. These include subspace identification approaches \cite{VanOverschee1991,Viberg1995}, and maximum likelihood estimates using either gradient-based algorithms (see e.g. \cite{Cappe2005,Sarkka2013}) or Expectation-Maximisation (EM) \cite{Shumway1982,Digalakis1993,Ghahramani1996,Gibson2005,Li2009}.

@@

The Gibbs sampler approach developed in \cite{Wills2012} is effective on the basic flavour of linear Gaussian state space models. However, it is not able to handle the case in which the transition model is degenerate, i.e. when the transition covariance matrix is not full rank. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. In addition, reversible jump moves \cite{Green1995,Green2009} are introduced which enable the sampler to learn explicitly the rank of the transition covariance matrix.

@@

The paper is structured as follows. In section~\ref{sec:linear_gaussian_models} we review the Gibbs sampler for basic linear Gaussian state space systems, focusing on the transition model. In section~\ref{sec:degenerate_transition_models} we describe modifications for degenerate transitions, and in section~\ref{sec:rank_learning} we show how the rank of the transition covariance may be learned. Simulations on a toy model, and a motion capture application are presented in section~\ref{sec:simulations} with discussion and conclusions in section~\ref{sec:conclusions}.

@@

The system is parameterised by the four matrices $\left\{\lgtm, \lgom, \lgtv, \lgov\right\}$. In this paper we focus on learning the transition model, i.e. $\lgtm$ and $\lgtv$, and for the most part treat $\lgom$ and $\lgov$ as fixed and known. Conditional on these matrices, state inference may be carried out analytically. Posterior filtering and smoothing densities may be calculated using the Kalman filter \cite{Kalman1960} and Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965}. Furthermore, the Kalman filter may also be used to evaluate the marginal likelihood $\den(\ob{1:\timax}|\lgtm, \lgtv)$, and it is possible to draw posterior samples from the state posterior $\den(\ls{1:\timax}|\ob{1:\timax}, \lgtm, \lgtv)$ using the forward-filtering-backward-sampling method \cite{Chib1996}. The purpose of a learning algorithm is to estimate unknown values of $\left\{\lgtm, \lgtv\right\}$ from a sequence of observations. Within the Bayesian framework this means calculating or approximating the posterior distribution $\den(\lgtm, \lgtv|\ob{1:\timax})$, which we can achieve using MCMC.

For Bayesian learning, we need to define prior distributions for the unknown system matrices $\lgtm$ and $\lgtv$. Although these should be selected to reflect prior belief about the system in question, there is substantial benefit in using conjugate priors, as these enable us to use Gibbs sampling moves. Typically, we will not have strong prior beliefs about the parameters, so it is reasonable to use an uninformative conjugate prior. However, if we do have prior knowledge to take into account which means that a conjugate prior is not appropriate, then we can treat each Gibbs move as a proposal in a Metropolis-Hastings scheme and use an accept/reject stage to account for the difference between the true prior and the conjugate prior used for the sampling \cite{Wills2012}.

@@

A commonly encountered variation on the basic linear Gaussian state space model is as follows,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \lgdm \tn{\ti} \label{eq:degenerate_transition} \\
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\idmat}      ,
\end{IEEEeqnarray}
%
in which $\tn{\ti}$ as dimension $\rk$, and $\lgdm$ is $\lsd\times\rk$. This is almost equivalent to the previous case, with $\lgtv=\lgdm\lgdm\tr$. However, whereas before $\lgtv$ was required to be positive definite, this is no longer the case. When this occurs, the conventional transition density $\den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgdm)$ is not defined, and the sampling operations underlying our Gibbs sampler are no longer valid. Degenerate models such as these arise in tracking applications (see \cite{Maskell2004} for a discussion of this phenomenon); when non-Markovian models, such as autoregressive process, are converted to state space form; and when it is natural to parameterise a system with more latent states than there are degrees of freedom (see section~\ref{sec:mocap}). In this section, we introduce suitable parameterisations for such degenerate transition models, and show how MCMC moves may be implemented for learning. Here we treat $\rk$ as fixed and known; in the following section mechanisms will be introduced to learn this too.

The matrix $\lgdm$ is not uniquely identifiable from the state sequence, because the distribution of $\lgdm\tn{\ti}$ is identical to that of $\lgdm \Xi \tn{\ti}$ where $\Xi$ is an arbitrary orthogonal matrix. This ambiguity does not affect $\lgtv$, since,
%
\begin{IEEEeqnarray}{rClCl}
 (\lgdm \Xi)(\lgdm \Xi)\tr &=& \lgdm\lgdm\tr &=& \lgtv     .
\end{IEEEeqnarray}
%
For this reason, we treat $\lgtv$ as fundamental, as before. $\lgdm$ may be defined to be any matrix square root of $\lgtv$ (e.g. the Cholesky factor or the principal matrix square root). In this new setting, $\lgtv$ is a positive semi-definite matrix with rank $\rk$.

@@

There is an obvious way to avoid being constrained to a particular subspace by the sampled state trajectory; simply return to targeting the marginal posterior $\postden(\lgtm, \lgtv|\ob{1:\timax})$ and fall back to using a generic Metropolis-Hastings approach. However, with a complex and high dimensional parameter space, designing efficient proposal distributions for Metropolis-Hastings is difficult. We would prefer to attain the automated, fast-mixing properties displayed by the Gibbs sampler on the full rank model. We will show how a Gibbs kernel may still be applied to sample a component of the parameters, with the remainder handled using Metropolis-Hastings.

@@

Note that we can find a set of $\lsd-\rk$ additional vectors orthonormal to $\tvvec$ using the Gram-Schmidt procedure. We denote the matrix of these vectors $\tvvecorth$. This matrix is not unique in general.

A further factorisation will also be useful, which we will refer to as the Givens decomposition (on account of its method of computation),
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition} %&=& \begin{bmatrix}\tvrot & \tvrotorth\end{bmatrix} \begin{bmatrix}\tvfull & \zmat \\ \zmat & \zmat\end{bmatrix} \begin{bmatrix}\tvrot\tr \\ \tvrotorth\tr\end{bmatrix}
\end{IEEEeqnarray}
%
where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors subject to certain constraints. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. See appendix~\ref{app:givens-factorisation} for details.

@@

An appropriate prior distribution for the singular transition covariance matrix may be defined by setting the degrees of freedom of an inverse Wishart distribution equal to the desired rank, $\priordof=\rk$. This results in a singular inverse Wishart distribution over the rank-$\rk$ positive semi-definite matrices \cite{Diaz-Garcia2006},
%
\begin{IEEEeqnarray}{rCl}
 \lgtv|\rk &\sim& \iwishartdist{\rk}{\priorscalematrix}     .
\end{IEEEeqnarray}
%
The density (with respect to the Lesbegue measure on the distinct elements) is,
%
\begin{IEEEeqnarray}{rCl}
 \den(\lgtv|\rk) &=& \frac{ \determ{\priorscalematrix}^{\frac{\rk}{2}} }{ 2^{\half\rk\lsd} \pi^{\half\rk(\lsd-\rk)} \gammafun[r]\left(\frac{\rk}{2}\right) } \nonumber \\
 & & \quad \times \determ{\tvval}^{-\half(3\lsd-\rk+1)} \exp\left( -\half \lgtv\pinv \priorscalematrix \right)    ,
\end{IEEEeqnarray}
%
in which $\lgtv\pinv$ denotes the Moore-Penrose pseudoinverse of $\lgtv$.

The matrix normal component of the prior is more problematic. Although the distribution is still well-defined, the fact that $\lgtv$ is singular means that the space of possible values is restricted to a subspace of $\reals^{\lsd\times\lsd}$, placing an unwanted constraint on our model. (See \cite{Diaz-Garcia2006} for a discussion of the singular matrix normal distribution.)

@@

Using this prior, we need a mechanism for choosing $\priorextratpval$. These extra eigenvalues control the rate at which probability decays as the matrix moves away from the subspace defined by $\lgtv$. Since $\tvvecorth$ can be arbitrarily rotated, it will generally be most suitable to use $\priorextratpval=\priortypval\idmat$ where $\priortypval$ is a prior estimate of a ``typical'' eigenvalue. Alternatively, $\priortypval$ could be made dependent on $\tvval$; for example it could be set to the average (arithmetic or geometric) of the current eigenvalues. In practice, uninformative priors will often be used. If $\priorcolumnvariance\to\infty$, then the effect of $\priortypval$ becomes negligible.

@@

As discussed, standard Gibbs sampling using $\postden(\lgtm, \lgtv|\ls{1:\timax})$ does not allow the Markov chain to fully explore the parameter space because of the constraint that each transition $(\ls{\ti}-\lgtm\ls{\ti-1})$ must lie in the $\rk$-dimensional column space of $\lgdm$. Nevertheless, we can sample from this distribution exactly, and doing so allows the chain to efficiently explore values of the transition matrix and transition covariance within the permitted regions. These constrained Gibbs sampling steps rely on the Givens decomposition \eqref{eq:givens_decomposition}.

@@

The likelihood function for the entire state trajectory is thus,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{ \den(\ls{1:\timax}|\lgtm, \lgtv) \propto \prod_{t=2}^{\timax} \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) } \nonumber \\
 &\propto& \determ{\tvfull}^{-\half(\timax-1)} \exp\Bigg( -\half \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \nonumber \\
 && \qquad\qquad\qquad\qquad\qquad\qquad\qquad \tvfull\inv \tvrot\tr (\ls{\ti}-\lgtm\ls{\ti-1}) \Bigg) \nonumber \\
 %&=& \determ{\tvfull}^{-\half(\timax-1)} \exp\left(-\half\trace\left[ \tvfull\inv \tvrot\tr \sum_{t=2}^{\timax} (\ls{\ti}-\lgtm\ls{\ti-1})(\ls{\ti}-\lgtm\ls{\ti-1})\tr \tvrot \right]\right) \nonumber \\
 %&=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\left(-\half\trace\left[ \tvfull\inv \left( \tvrot\tr \lgtm \suffstats{1} \lgtm\tr \tvrot - \tvrot\tr \lgtm \suffstats{2}\tr \tvrot - \tvrot\tr \suffstats{2} \lgtm\tr \tvrot + \tvrot\tr \suffstats{3} \tvrot \right) \right]\right) \\
 &=& \determ{\tvfull}^{-\half\suffstats{0}} \exp\bigg(-\half\trace \big[ \tvfull\inv ( \lgtmrot \suffstats{1} \lgtmrot\tr - \lgtmrot \suffstats{2}\tr \tvrot \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad -\: \tvrot\tr \suffstats{2} \lgtmrot\tr + \tvrot\tr \suffstats{3} \tvrot ) \big] \bigg)
\end{IEEEeqnarray}
%
in which the sufficient statistics are as before and,
%
\begin{equation}
 \lgtmrot = \tvrot\tr \lgtm      .
\end{equation}

Writing all the subspace constraints concisely in a single equation,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \lgtm \LS{1:\timax-1} + \tvrot Z
\end{IEEEeqnarray}
%
where $Z \in \reals^{\rk\times(\timax-1)}$ and 
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& \begin{bmatrix} \ls{2} & \ls{3} & \hdots & \ls{\timax} \end{bmatrix}    .
\end{IEEEeqnarray}

@@

\section{Unknown Covariance Matrix Rank} \label{sec:rank_learning}

@@

\subsection{Priors with Unknown Rank}

We must take care in setting $\priorscalematrix$ in $\den(\lgtv)$. In general, the scale of an inverse Wishart random variable depends on both the scale matrix and the degrees of freedom. If the same $\priorscalematrix$ were used for all $\rk$ then the prior would favour different scales with different possible values of the rank. In order to make the prior scale invariant across $\rk$, we replace $\priorscalematrix$ with $\priorscalematrixrank{\rk} = \rk \priorscalematrixbase$. $\priorscalematrixbase$ sets the scale of $\lgtv$ globally.

We also introduce a prior distribution for the rank of $\lgtv$,
%
\begin{IEEEeqnarray}{rCl}
 \den(\rk) &=& \sum_{i=1}^{\lsd} \priorrankweight{i} \diracden{\rk}{i}     ,
\end{IEEEeqnarray}
%
where $\diracden{j}{i}$ is a probability point mass at $j=i$, and $\sum_i \priorrankweight{i} = 1$. The target distribution is now,
%
\begin{IEEEeqnarray}{rCl}
 \postden(\lgtm, \lgtv, \rk) &\propto& \den(\ob{1:\timax}|\lgtm, \lgtv) \den(\lgtm, \lgtv|\rk) \den(\rk)    .
\end{IEEEeqnarray}

\subsection{Reversible Jump for Rank Learning}

@@

For the sake of simplicity we suggest sampling the new eigenvalue and eigenvector independently from uniform distributions. For the eigenvalue,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) = \begin{cases}
                                          \frac{1}{\eval{\rk}} & 0 < \eval{\rk+1} < \eval{\rk} \\
                                          0 & \text{ otherwise}    .
                                         \end{cases}
\end{IEEEeqnarray}
%
For the eigenvector,
%
\begin{IEEEeqnarray}{rCl}
 \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) = \begin{cases}
                                          \frac{1}{C_{\rk+1}} & \evec{\rk+1} \in \evecset{\rk+1} \\
                                          0 & \text{ otherwise}    .
                                         \end{cases}
\end{IEEEeqnarray}
%
$C_{\rk+1}$ is the volume of $\evecset{\rk+1}$, which is half of a $(d-r-1)$-sphere (half because of the requirement that the sign of the first element be positive) \cite{Muirhead1982}. Hence,
%
\begin{IEEEeqnarray}{rCl}
 C_{\rk+1} &=& \frac{\pi^{\half(\lsd-\rk)}}{\gammafun\left(\frac{\lsd-\rk}{2}\right)}     .
\end{IEEEeqnarray}
%
In order to generate such a vector, first draw a $(\lsd-\rk)$-dimensional sample from $\normaldist{\zmat}{\idmat}$, normalise it, and then multiply by $\tvvecorth$.

The acceptance probability for this move and its reverse respectively are \cite{Green1995},
%
\begin{IEEEeqnarray}{rCl}
 \mhap(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) &=& \min\left\{1, \beta \right\} \\
 \mhap(\lgtv\mcnew\to\lgtv,\eval{\rk+1},\evec{\rk+1}) &=& \min\left\{1, \beta\inv \right\} 
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \beta &=& \frac{ \postden(\lgtm, \lgtv\mcnew, \rk+1) \jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew) }{ \postden(\lgtm,\lgtv,\rk) \ppslden{\eval{}}(\eval{\rk+1}|\lgtv) \ppslden{\evec{}}(\evec{\rk+1}|\lgtv) } \nonumber \\
 & & \qquad\qquad\qquad\qquad\qquad \times \frac{P(\rk+1\to\rk)}{P(\rk\to\rk+1)}     ,
\end{IEEEeqnarray}
%
where $\jac(\lgtv,\eval{\rk+1},\evec{\rk+1}\to\lgtv\mcnew)$ is the Jacobian of the transformation and $P(\rk\to\rk+1)$ is the probability of performing a move which changes the rank from $\rk$ to $\rk+1$.

@@

\section{Discussion and Conclusions} \label{sec:conclusions}
